---
title: "Programming with fixed-points"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{fixed-point-programming}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup, warning=FALSE, message=FALSE, echo=FALSE}
library(recursr)
```

# Preface

When I was exploring catamorphisms, I originally implemented these using the functions `Fix()` and `unFix()` which behaved like the Haskell counterparts. In contrast, the `recursion-schemes` package implements `project()` and `embed()` functions together, with "base" functors for e.g. lists, with the idea being that the "base" functor represents one layer of recursion. I don't think these names are particularly intuitive, but `project()` is equivalent to `unFix()` as it peels back a layer of recursion, and `embed()` is the equivalent of `Fix()` as applies a layer of recursion. 

I originally followed Mark Seemann's [blog](https://blog.ploeh.dk/2019/05/13/peano-catamorphism/), which in turn uses `Fix` as in Bartosz Milewski's [article on F-Algebras](https://bartoszmilewski.com/2017/02/28/f-algebras/). Furthermore, a [Haskell Wikibook section](https://en.wikibooks.org/wiki/Haskell/Fix_and_recursion) explores Haskell's `fix` and recursion.

# Fixed points

Cheating a bit—`Fix()` just wraps another layer of recursion, it does not have a recursive type signature like Haskell's `fix`—I defined the `Fix()` and `unFix()` functions as follows.

```{r defineFixedPoint}
Fix <- function (expr) {
    res         = list (type = "Fix", content = expr)
    class (res) = append (class (res), "fixed.point")
    return (res)
}

unFix <- function (fx) {
    if ("fixed.point" %in% class (fx))
        return (fx$content)
}

is.fixed.point <- function (fx) {
    return ("fixed.point" %in% class (fx))
}
```

# Redefining `cata()` and `ana()`

Let's define `cata()` and `ana()` in terms of `unFix()` and `Fix()` (they just slot in).

```{r defineCataAnaFix}
cata_ <- function (alg, expr)
    unFix (expr) |> fmap (\(x) cata_ (alg, x)) |> alg ()

ana_ <-  function (coalg, expr)
    coalg (expr) |> fmap (\(x) ana_ (coalg, x)) |> Fix ()
```

# Defining Peano natural numbers

The Peano zero is defined, as is the function `Succ()` to wrap it. These are all very straightforward.

```{r definePeanoNumbers}
Zero <- function () {
    res         = list (type = "Zero")
    class (res) = "Natural"
    return (res)
}
Succ <- function (x) {
    res         = list (type = "Succ", content = x)
    class (res) = "Natural"
    return (res)
}

is.Zero <- function (n) {
    if ("Natural" %in% class (n))
        return (n$type == "Zero")
    else return (FALSE)
}

is.Succ <- function (n) {
    if ("Natural" %in% class (n))
        return (n$type == "Succ")
    else return (FALSE)
}

fmap.Natural <- function (st, f) {
    if (is.Succ (st)) {
        x  = f (st$content)
        return (Succ (x))
    }
    else if (is.Zero (st))
        return (st)
}
```

# Test data

Now, define some test data, numbers 0-3.

```{r defineTestData}
zero  = Fix (Zero ())
one   = Fix (Succ (zero))
two   = Fix (Succ (one))
three = Fix (Succ (two))
```

# Simple recursive functions

Given that  we have the catamorphism, we can just go ahead and define functions which call `cata()` in order to pretty-print and tally up values.

```{r defineSampleFunctions}
printNats <- function (st) {
    pn = function (x) {
        if (is.Zero (x))
            return ("Zero")
        else if (is.Succ (x)) {
            exprStr = paste ("Succeeds (", x$content, ')', sep='')
            return (exprStr)
        }
    }
    res = cata_ (pn, st)
    return (res)
}

tallyNats <- function (st) {
    tn = function (x) {
        if (is.Zero (x))
            return (0)
        else if (is.Succ (x))
            return (x$content + 1)
    }
    res = cata_ (tn, st)
    return (res)
}
```

Example output: 

```{r evalTestData}
printNats (zero)
tallyNats (zero)

printNats (one)
tallyNats (one)

printNats (two)
tallyNats (two)

printNats (three)
tallyNats (three)
```

# More recursive functions

Following Seemann's blog post linked earlier, we can define additional functions:

  - `natF()`, which applies a function `f` as many depths as the structure;
  - `incr()`, which increments a given Peano representations by one;
  - `add()`, which adds two Peano representations;
  - `mult()`, which multiplies two Peano representations;
  - `toNum()`, which "expands" a given Peano expression into an integer representation;
  - `fromNum()`, which "compresses" a given whole number into a Peano representation.
  
```{r defineMoreFunctions}
natF = function (z, f, expr) {
    alg = function (x) {
        if (is.Zero (x))
            return (z)
        else if (is.Succ (x)) {
            pre = x$content
            return (f (pre))
        }
    }
    return (cata_ (alg, expr))
}

incr <- function (expr)
    return (natF (one, \(x) Fix (Succ (x)), expr))

add <- function (ex, ey)
    return (natF (ey, incr, ex))

mult <- function (ex, ey)
    return (natF (zero, \(x) add (ey, x), ex))

toNum <- function (expr) {
    expand = function (st) {
        if (is.Zero (st))
            return (0)
        else if (is.Succ (st)) {
            return (st$content + 1)
        }
    }
    res = cata_ (expand, expr)
    return (res)
}

fromNum <- function (n) {
    compress = function (x) {
        if (x == 0)
            return (Zero ())
        else if (x > 0)
            return (Succ (x-1))
    }
    res = ana_ (compress, n)
    return (res)
}
```

Example output:
  
```{r evalMoreFunctions}
fromNum (4)  |> printNats ()
fromNum (32) |> tallyNats ()
incr (fromNum (4))               |> tallyNats ()
add  (fromNum (4), fromNum (2))  |> tallyNats ()
mult (fromNum (4), fromNum (12)) |> tallyNats ()
```

# Final words

The main draw-back to this approach is that it requires defining recursive 
fixed point variables, whereas the approach taken by the `recursion-schemes` 
package with the "base" functors allows for working directly on the structures 
in question. However, being able to see how `unFix()` and `Fix()` relate to the
`project()` and `embed()` functions is useful as a learning exercise.
