---
title: "Morphisms for a list"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{list-morphisms}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup, message=F, warning=F}
library(recursr)
```

# Folds
## Catamorphism

The catamorphism is a universal abstraction for tearing down a functor structure.

With simple recursion is feasible to print a list as follows:

```{r p1pt1rec}
printListSimple <- function (lst) {
    if (length (lst) == 0)
        return ("nil")
    else {
        topNode  = lst[[1]]
        allNodes = lst[-1]

        return (paste (topNode, ": ", printListSimple (allNodes), sep=''))
    }
}

printListSimple (as.list (1:10))
```

This can be rewritten with a catamorphism:

```{r p1p2cata}
printListCata <- function (lst) {
    alg = function (l) {
        if (is.NilF (l))
            return ("nil") 
        else if (is.ConsF (l))
            return (paste (l$attr, ": ", l$carrier, sep='', collapse=''))
    }
    return (cata (alg, lst))
}

printListCata (as.list (1:10))
```

## Paramorphism

The paramorphism is a variant of the catamorphism, one which provides access to
the current sub-structure at any given point. The algebra needs to interpret
tuples in the carrier type.

The following function appends the size of the sub-structure to the end of each 
list element:

```{r p1p3para}
printIndent <- function (lst) {
    alg = function (l) {
        if (is.NilF (l))
            return (list())
        else {
            x  = l$attr
            xs = l$carrier
            xsa = first (xs)
            xsc = second (xs)

            indent = replicate (length (xsa), '>')
            pasted = paste (indent, sep='', collapse='')
            
            return (append (xsc
                          , paste (x, pasted, sep='')
                          , after=0))
        }
    }
    return (para (alg, lst))
}

printIndent (list (1,2,3,4))
```

## Histomorphism

The histomorphism goes further still by providing access to all previous values
in the recursion. These are stored in a structure provided by cofree, embedded
throughout the list base functor's carrier.

The following function skips over elements at even indices. This function is from https://jtobin.io/time-traveling-recursion.

```{r p1p4histo}
currElement <- function (l)
    return (l$attr)

peelBack <- function (l)
    return (l$carrier)

oddIndices <- function (lst) {
    alg = function (l) {
        if (is.NilF (l))
            return (list())
        else {
            h0  = currElement (l)
            cf0 = peelBack (l)
            xs0 = peelBack (cf0)

            if (is.NilF (xs0))
                return (list (h0))
            else {
                h1  = currElement (xs0)
                cf1 = peelBack (xs0)
                t1  = currElement (cf1)

                return (c(h0, t1))
            }      
        }
    }
    return (histo (alg, lst))
}

oddIndices (as.list (1:10))
```

# Unfolds
## Anamorphism

The anamorphism is the categorical dual of the catamorphism. It unfolds a 
structure into a list, rather than folding a list structure into something else,
i.e. it builds up a structure rather than tearing it down like the catamorphism
does.

The following functions are the equivalent of `as.list(unlist(recursive=T))`. 
They build up a simple list from a more complex structure, in this case a list 
of lists. Compare the lack of recursive calls in the latter, to the former.

```{r p2p1rec}
collapseListsSimple <- function (lst) {
    if (length (lst) == 0)
        return (list ())
    else {
        x  = lst[[1]]
        xs = lst[-1]
        
        if (is.list (x) && length (xs) == 0)
            collapseListsSimple (x)
        else if (!is.list (x) && length (xs) == 0)
            list (x)
        else if (is.list (x) && length (xs) > 0)
            c (collapseListsSimple (x), collapseListsSimple (xs))
        else if (!is.list (x) && length (xs) > 0)
            append (collapseListsSimple (xs), x, after=0)
    }
}

collapseListsSimple (list(list(1,2,3),list(4,5,6),list(list(7,8),9,list(10,11))))
```

```{r p2p2ana}
collapseListsAna <- function (lst) {
    coalg = function (x) {
        if (length (x) == 0)
            return (NilF())
        else {
            x_ = x[[1]]
            xs = x[-1]

            if (length (xs) == 0)
                 ConsF (x_, list())
            else if (length (xs) == 1)
                 ConsF (x_, xs[[1]])
            else ConsF (x_, xs)
        }
    }
    return (ana (coalg, lst))
}

collapseListsAna (list(list(1,2,3),list(4,5,6),list(list(7,8),9,list(10,11))))
```

## Apomorphism

The apomorphism is the categorical dual of the paramorphism. It allows the 
recursion to be terminated at any point.

The following function is an implementation of the tails function. This function
is designed returns a list of the tails of a given list as the apomorphism 
traverses the list structure.

Note the use of `Left()` and `Right()`. If a `Left` value is provided, the 
paramorphism will just return whatever value is in there, i.e. we've stopped
recursion. If a `Right` value is provided, recursion continues.

```{r p2p2apo}
tailsApo <- function (lst) {
    coalg = function (a) {
        projected = project (a)
        if (is.NilF (projected))
            ConsF (list (list ()), Left (list ()))
        else {
            xs = projected$carrier
            ConsF (list (a), Right (xs))
        }
    }
    apo (coalg, lst)
}

tailsApo (list (1,0,2,4,3))
```

It's not very useful here, but the function doesn't have to stop on the empty 
list. It can stop on arbitrary leading values, e.g. zero:

```{r p2p3apo}
tailsApoZero <- function (lst) {
    coalg = function (a) {
        projected = project (a)
        
        if (projected$attr == 0)
            ConsF (list (a), Left (list()))
        else if (is.NilF (projected))
            ConsF (list (list ()), Left (list ()))
        else {
            xs = projected$carrier
            ConsF (list (a), Right (xs))
        }
    }
    apo (coalg, lst)
}

tailsApoZero (list (1,0,2,4,3))
```

## Futumorphism

The futumorphism is the categorical dual of the histomorphism. Instead of access
to previously-computed answers, the futumorphism provides access to values yet 
to be computed.

An implementation of the exchange function is given. This example as usually
implemented swaps pairs. This example instead swaps sets of three as it shows
better how to access the 'future' values. R lacks the pattern matching which
Haskell has; as such the Haskell equivalent is much more succint.

```{r p2p3futu}
exchange <- function (lst) {    
    coalg = function (a) {
        projected = project (a)
        
        if (is.NilF (projected))
            return (NilF())
        else {
            x = projected$attr
            k = projected$carrier
            projected_ = project (k)
  
            if (is.NilF (projected_))
                return (ConsF (x, liftF (NilF())))
            else {
                x_ = projected_$attr
                k_ = projected_$carrier
                projected__ = project (k_)
                
                if (is.NilF (projected__))
                    return (ConsF (x_, liftF (ConsF (x, k_))))
                else {
                    x__ = projected__$attr
                    k__ = projected__$carrier
                    
                    return (ConsF (x__, Free (ConsF (x_, liftF (ConsF (x, k__))))))
                }
            }
        }
    }
    futu (coalg, lst)
}

exchange (as.list (1:9))
```
