---
title: "Morphisms for a tree"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{tree-morphisms}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup, warning=FALSE, message=FALSE, echo=FALSE}
library(recursr)
```

# Preface

This vignette is a tutorial which walks through implementation of base functors
for an existing tree object (`Node`, from `data.tree`), followed by implementing
a custom `Tree` type which behaves similarly. 

For both types, the tutorial guides building some functions which make use of 
the morphisms included in the package. However, as will become clear, writing
unfolds (`apo`, `para`, `futu`) for the `data.tree` `Node` is problematic. 
A problem does arise in that when adding child nodes which are named the same as
an existing node in the tree, the behaviour is unpredictable.

Sample code which produces this behaviour will be included in the tutorial,
and will lead directly on to a custom `Tree` type which, while not as 
fully-featured as the functionality of `data.tree`, does behave predictably with
duplicate nodes.

# The `data.tree` `Node` type
## Defining flat-map

```{r initNode}
library (data.tree)

fmap.Node <- function (tr, f) {
    working = Clone (tr, attributes=F)

    traversal = Traverse (working)
    Do (traversal, function(node) node$name <- f (node$name))
    
    return (working)
}
```

Let's start by implementing flat-map for the `data.tree` `Node` object.
Confusingly, referencing a `data.tree` object with a new variable also
affects the original object. We need to clone the object, which is not very 
efficient, but fine for this example, which always returns a new object without
affecting the original.

## Defining the "base" functor

```{r initNodeF}
NodeF <- function (r, l) {
    res         = list (type = "NodeF", root=r, forest=l)
    class (res) = "NodeF"
    return (res)
}
is.NodeF <- function (trf) {
    return ("NodeF" %in% class (l))
}
fmap.NodeF <- function (nf, f) {
    if (length (nf$forest) == 0)
        return (NodeF (nf$root, list()))
    else
        return (NodeF (nf$root, fmap (nf$forest, f)))
}
```

The "base" functor of a `Node` object has a label/attribute type called root 
and a carrier type called forest. 
The flat-map function acts on the carrier type, *unless* the forest is
empty (we've hit a 'leaf') in which case it acts on the attribute type.

## Translating between the functor and its "base" functor

We need two functions which unwrap and wrap a level of recursion, respectively.
The `project` function will take a `Node` and "project" one layer of it onto the
"base" functor, and the `embed` function will take a `Node` "base" functor and
"embed" one layer of it in a tree structure.

The `project` function for a `Node` object is as follows:

```{r projectEmbed}
project.Node <- function (node)
    return (NodeF (node$name, node$children))
```

In contrast, the writing an `embed()` function ends up being problematic 
because of the behaviour of `Node$AddChildNode` with duplicated nodes, 
but assuming that no duplicates are extant, the following function works:

```{r embedNodeF}
embed.NodeF <- function (nf) {
    if (length (nf$forest) == 0)
        return (Node$new (nf$root))
    else {
        staging = Node$new (nf$root)
        lapply (nf$forest, staging$AddChildNode)
        return (staging)
    }
}
```

## Conversion to a list of lists and vice versa

The `data.tree` package includes the functions `ToListSimple` and
`FromListSimple`. 
This example uses the `acme` example in the package vignette as it contains no 
duplicate nodes, converts it to a list of lists, and then 
demonstrates use of cata and ana to convert between a list of lists and our own
Tree type. 

Note that these functions assume that there is only one label as in the vignette
example. In `data.tree`, it is actually possible to have an arbitrary number of
labels, although they aren't displayed by the default printing function.

Here is some sample data as given by the `data.tree` [introduction](https://cran.r-project.org/web/packages/data.tree/vignettes/data.tree.html):

```{r genAcme}
library (data.tree)

acme <- Node$new("Acme Inc.")
  accounting <- acme$AddChild("Accounting")
    software <- accounting$AddChild("New Software")
    standards <- accounting$AddChild("New Accounting Standards")
  research <- acme$AddChild("Research")
    newProductLine <- research$AddChild("New Product Line")
    newLabs <- research$AddChild("New Labs")
  it <- acme$AddChild("IT")
    outsource <- it$AddChild("Outsource")
    agile <- it$AddChild("Go agile")
    goToR <- it$AddChild("Switch to R")
    
acme
```

Now that we have a list of lists to work with, we can use `ana` to build up a 
`Node` structure, and `cata` to tear it back down:

```{r fnl}
fromNestedList <- function (lst) {
    fromNested = function (nl) {
        NodeF (nl$name, nl[-1])
    }
    ana (fromNested, lst)
}

toNestedList <- function (node) {
    toNested = function (nf) {
        branch = list (name = nf$root)
        return (append (branch, nf$forest))
    }
    cata (toNested, node)
}
```

We can now combine the two to get the same result as we put in:

```{r combineRes}
acme |> ToListSimple () |> fromNestedList () |> toNestedList () |> FromListSimple ()
```

## Pretty-printing trees

The following example is based on the one from [Stack Overflow](https://stackoverflow.com/questions/12556469/nicely-printing-showing-a-binary-tree-in-haskell/19442407#19442407), although it generalises to non-binary trees.

```{r naiveDrawTree}
drawTree <- function (tree) {
    pad = function (body, final, l) {
        if (length (l) == 0)
            return (list())
        else if (length (l) == 1)
            return (list (paste (body, l[[1]], sep='')))
        else {
            firstPortion = l[[1]]
            lastPortion  = l[2:length(l)]

            res = append (list (paste (body, firstPortion, collapse='', sep=''))
                        , lapply (lastPortion, \(x) paste (final, x, sep='')))
            return (res)
        }
    }

    prettyPrintHelper <- function (br) {
        if (length (br$forest) == 0)
            return (list (br$root))
        else
            return (append (list (br$root), prettyPrintForest (br)))
    }
    prettyPrintForest <- function (br) {
        forest = br$forest
        forlen = length (forest)

        if (forlen == 1)
            return (pad ("┗━ ", "   ", prettyPrintHelper (forest[[1]])))
        else {
            padding1 = lapply (forest[1:(forlen-1)]
                             , \(x) pad ("┣━ ", "┃  ", prettyPrintHelper (x)))
            padding2 = pad ("┗━ ", "   ", prettyPrintHelper (forest[[forlen]]))

            return (append (padding1, padding2))
        }
    }
    cat (paste (as.list (unlist (prettyPrintHelper (tree), recursive=F))
              , collapse="\n", sep='')
       , '\n')
}
```

The main advantage of using `cata`, as follows, is being able to reduce 
recursive calls of `prettyPrintForest()` and `prettyPrintHelper()`, which,
confusingly, call each other. 
Otherwise, the function which follows is still fairly complicated. 
Given that `apo` gives access to level information, and `histo` gives access to 
past results, it should be feasible to write a function which makes use of these.

```{r betterDrawTree}
drawTreeCata <- function (tree) {
    pad = function (body, final, l) {
        if (length (l) == 0)
            return (list())
        else if (length (l) == 1)
            return (list (paste (body, l[[1]], sep='')))
        else
            return (append (list (paste (body, l[[1]], collapse='', sep=''))
                          , lapply (l[2:length(l)], \(x) paste (final, x, sep=''))))
    }
    prettyPrint <- function (br) {
        if (length (br$forest) == 0)
            return (list (br$root))
        else if (length (br$forest) == 1)
            return (append (pad ("┗━ ", "   ", br$forest), br$root, after = 0))
        else {
            forest = br$forest
            forlen = length (forest)
            
            padding1 = lapply (forest[1:forlen-1], \(x) pad ("┣━ ", "┃  ", x))
            padding2 = pad ("┗━ ", "   ", forest[[forlen]])
            padding = append (padding1, padding2)
            
            return (append (padding, br$root, after = 0))
        }
    }
    res = cata (prettyPrint, tree)
    cat (paste (purrr::flatten (res), collapse='\n', sep=''), '\n')
}

acme |> ToListSimple () |> fromNestedList () |> drawTreeCata ()
```

## Tripping up `data.tree`

While the above examples work fine, it's fairly trivial to trip `data.tree` up.
The expected behaviour is to add another, identical sub-tree, 
called "rabbit family", but the sub-trees end up getting mixed, or lost.

```{r genDups}
dups <- Node$new("rabbit warren")
  family1 <- dups$AddChild ("rabbit family")
    rabbit1.1 <- family1$AddChild ("rabbit matriarch")
    rabbit1.2 <- family1$AddChild ("rabbit baby")
    rabbit1.3 <- family1$AddChild ("rabbit engineer")
  family2 <- dups$AddChild ("rabbit family")
    rabbit2.1 <- family2$AddChild ("rabbit matriarch")
    rabbit2.2 <- family2$AddChild ("rabbit scout")
    rabbit2.3 <- family2$AddChild ("rabbit professor")
  family3 <- dups$AddChild ("solitary pair of hares")
    hare1 <- family3$AddChild ("solitary hare #1")
  family4 <- dups$AddChild ("solitary hare #2")
  family5 <- dups$AddChild ("solitary hare #2")

dups
```

# A custom `Tree` type
## Defining a custom `Tree` type

Since we're not using an existing type, we need to define functions which generate the new S3 type `Tree`, and a flat-map implementation.

```{r custTree}
Tree <- function (r, l) {
    res         = list (type = "Tree", root=r, forest=l)
    class (res) = "Tree"
    return (res)
}
is.Tree <- function (x) {
    return ("Tree" %in% class (x))
}

fmap.Tree <- function (tr, f) {
    if (length (tr$forest) == 0)
        return (Tree (f (tr$root), list()))
    else {
        newForest = fmap (tr$forest, \(x) fmap.Tree (x, f))
        return (Tree (f (tr$root), newForest))
    }
}
```

## Defining the "base" functor

Now, define `TreeF`, the "base" functor of `Tree`. This is near-identical to `NodeF`, which will come in handy later.

```{r custTreeF}
TreeF <- function (r, l) {
    res         = list (type = "TreeF", root=r, forest=l)
    class (res) = "TreeF"
    return (res)
}

is.TreeF <- function (trf) {
    return ("TreeF" %in% class (l))
}

fmap.TreeF <- function (trf, f) {
    if (length (trf$forest) == 0)
        return (TreeF (trf$root, list()))
    else
        return (TreeF (trf$root, fmap (trf$forest, f)))
}
```

## Translating between the functor and its "base" functor

Now, define `project()` and `embed()` for `TreeF` and `Tree`.

```{r custTreeProjEmbed}
project.Tree <- function (tr) {
    return (TreeF (tr$root, tr$forest))
}
embed.TreeF <- function (trf) {
    return (Tree (trf$root, trf$forest))
}
```

## Redefining the `fromNestedList()` function

Recall that the "base" functor of our new `Tree` type is near-identical to the
"base" functor of the `data.tree` `Node` type. 
Both `NodeF` and `TreeF` are S3 objects which are fundamentally a list with 
`$carrier` and `$forest` attributes. 
Therefore, it should be sufficient just to redefine `fromNestedList()`, the only 
change of which is returning `TreeF` as opposed to `NodeF`. 
Indeed, it should be possible to call `drawTreeCata()` directly on a `Tree` 
object.

```{r fnl2}
fromNestedList_ <- function (lst) {
    fromNested = function (nl) {
        TreeF (nl$name, nl[-1])
    }
    ana (fromNested, lst)
}

acme |> ToListSimple () |> fromNestedList_ () |> drawTreeCata ()
```

## Implementing a monad instance

As a bonus, we can also define a monad instance for the new `Tree` type.

```{r custTreeBind}
bind.Tree <- function (tr, f) {
    x   = tr$root
    txs = tr$forest
    
    k = f(x)
    y   = k$root
    tys = k$forest
    p = lapply (txs, \(a) a %>>=% f)
    return (Tree (y, c(tys, p)))
}
```

Let's check the monad laws, which can be summarised as follows:

1. Left identity:
```
    return >=> h ≡ h
```
2. Right identity:
```
    f >=> return ≡ f
```
3. Associativity:
```
    (f >=> g) >=> h ≡ f >=> (g >=> h)
```

First, define the equivalent of Haskell's `return()`, which I've called `escape`:

```{r custTreeMonadLaws}
escape.Tree <- function (x)
    return (Tree (x, list()))

f <- function (x)
    return (Tree (x+2, list (escape.Tree (x+3), escape.Tree (x+4))))
g <- function (x)
    return (Tree (x+4, list ()))
```

Now, implement the laws above in code:

```{r custTreeEvalMonadLaws}
leftIdentityLHS <- escape.Tree (3) %>>=% f
leftIdentityRHS <- f (3)

rightIdentityLHS <- f (3) %>>=% escape.Tree
rightIdentityRHS <- f (3)

associativityTestTree <- Tree (3, list (
                           Tree (4, list())
                         , Tree (5, list(Tree (6, list()), Tree (7, list())))
                         ))
associativityLHS <- (associativityTestTree %>>=% f) %>>=% g
associativityRHS <- associativityTestTree %>>=% (\(x) f(x) %>>=% g)
```

Evaluate equivalence by drawing the trees to the terminal:

```{r custTreeLIdent}
leftIdentityLHS |> drawTreeCata ()
leftIdentityRHS |> drawTreeCata ()
```
```{r custTreeRIdent}
rightIdentityLHS |> drawTreeCata ()
rightIdentityRHS |> drawTreeCata ()
```
```{r custTreeAssoc}
associativityLHS |> drawTreeCata ()
associativityRHS |> drawTreeCata ()
```
It looks like our type satisfies the monad laws. However, implementing this as a
unit test, as for the list type, is non-trivial, as R has no way to know that
two `Tree` objects are equal. Writing a function to compare the structure / node
values could enable formally testing equality of the above LHS/RHS trees.

Anyway, implementing bind enables neat operations like the following:

```{r custTreeBindTest}
acme_ <- acme |> ToListSimple () |> fromNestedList_ ()

treeToUpper <- purrr::compose (escape.Tree, toupper)

acme_ %>>=% treeToUpper |> drawTreeCata ()
```

